### 2016-02-07 mcmc.tex
eradicated four HOGGs

% to-do
% -----
% - Kill all issues in this list, below:
% - Deal with all mentions of ``DFM'' or ``HOGG'' and ``...''
% - Should we cite the Brooks book or just chapters within it?  I think the latter!
% - Should we make an \acronym{} environment and use it religiously?  I think yes.
% - Do we need to explain the variance on a sampling-based estimator of an integral?  Right now this is just in an end-note (or will be when I write it).  Should it be in the main text?  Probably!
% - mention ESJD
% - mention Gelman--Rubin statistic
% - mention JAGS and STAN in methods section
% - follow up the idea that burn-in doesn't exist at all!  It does for emcee.
% - when we say ``below'' or ``above'' also give a Section Number.
% - need to mention case where there is an MC integral INSIDE the likelihood function
% - need to mention testing the likelihood function
% - would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
% - send to friendlies for comments
%   - Gelman
%   - Goodman
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
% - post on arXiv

% style notes
% -----------
% - These documents are ``Chapters'', no?  Use \documentname.
% - [LaTeX] newline after every full stop for proper git diffing
% - [LaTeX] eqnarray not equation
% - pdf not PDF
% - is it ``a MCMC sampling'' or ``an MCMC sampling''?
% - \note{} right after word if you are endnoting a phrase,
%   but after the full stop if endnoting the sentence or paragraph.
% - always $\pars$ never $x$; $K$ samples $\pars_k$ and $\pars$ space is $D$-dimensional
% - should we use the word ``walker'' everywhere and do we?

@@

\newcommand{\Var}{\mathrm{Var}}
\newcommand{\data}{D}
\newcommand{\pars}{\theta}
\newcommand{\best}{{\mathrm{(best)}}}
\newcommand{\better}{{\mathrm{(better)}}}

@@

\begin{abstract}
Markov Chain Monte Carlo (MCMC) methods for sampling probability
distribution functions (combined with abundant computational resources)
have transformed the sciences.
Here we give a brief overview of the most basic MCMC method and then turn to
practical advice for the use of MCMC in real inference problems.
We give advice on method choice, tuning for performance,
methods for initialization, tests of convergence, troubleshooting, and use of the chain
output to produce or report parameter estimates with associated uncertainties.
\end{abstract}

@@

Markov Chain Monte Carlo (MCMC) methods are methods for sampling
probability distribution functions or probability density functions (pdfs).
These pdfs may be either probability mass functions on a discrete
space, or probability densities on a continuous space, though we will
concentrate on the latter in this \documentname.
MCMC methods don't require that you have a full analytic description of the
properly normalized pdf for sampling to proceed; they only require
that you be able to compute ratios of the pdf at pairs of locations.
This makes MCMC methods ideal for sampling \emph{posterior
  pdfs} in probabilistic inferences:

@@

In addition to this normalization-insensitive property of MCMC, in its
simplest forms it can be run without computing any derivatives or
integrals of the function, and (as we will show below) in its simplest
forms it is \emph{extremely easy to implement}.
For all these reasons, MCMC is ideal for sampling posterior pdfs in
the real situations in which scientists find themselves.

Say you are in this situation:
You have a huge blob of data $\data$ (think of this as a vector or
list or heterogeneous collection of observations\note{In general in
  this \documentname\ we will treat any variable as being capable of
  containing not just a value but some huge arbitrarily structured set
  of values in matrix, vector, or an ordered list of any set of such objects,
  or perhaps some even more complex form.}).
You also have a model sophisticated enough---a probabilistic,
generative model, if you will\note{For more discussion
  of our definition of a \emph{model}, see [some other as-yet-unwritten \documentname] in
  this series.
  Briefly, a ``model'' for us is a likelihood function (a pdf for the data given
  model parameters), and a prior pdf over the parameters.
  Because, under this definition, the model can always generate (by
  sampling, say) parameters and parameters can generate (again by
  sampling, say) data, the model is effectively (or actually, if you
  are a true subjective Bayesian) a probability distribution (pdf)
  over all possible data.%
}---that, given a
setting of a huge blob of parameters\note{Again, the parameter
  blob $\pars$ should also be thought of as an immense, arbitrarily
  structured set or list of parameters, possibly even with no fixed
  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We will
  bash the terminology ``likelihood'' in another, as-yet-unwritten, \documentname\ in this
  series.  Techically, this thing $p(\data\given\pars)$
  is only properly a likelihood function when
  we are thinking of the data $\data$ as being fixed, and the parameters $\pars$
  as being permitted
  to vary.}) $p(\data\given\pars)$.
Furthermore, say also that you can write down some kind of informative or
vague prior pdf $p(\pars)$ for the parameter blob $\pars$.
If all these things are true, then---even if you can't compute
anything else---in principle a trivial-to-implement MCMC can give you
a fair sampling of the posterior pdf.
That is, you can run MCMC
(for a very long time---see \sectionname~\ref{sec:convergence} for how long)
and you will be left with a set of $K$ parameter-blob settings $\pars_k$ such that the
full set $\setof{\pars_k}_{k=1}^K$ constitutes a fair sampling from
the posterior pdf $p(\pars\given\data$).
% ************* BJB
% Notions of "sampling" and "Monte Carlo" are interesting. You might like to
% refer the interested reader to the paper "Monte Carlo is Fundamentally
% Unsound" by Tony O'Hagan.

All that said, and adhering to the traditions of the \project{Data
  Analysis Recipes} project\note{Every \documentname\ in the \project{Data
    Analysis Recipes} series begins with a rant in which we argue that
  most uses of the methods in question are not appropriate!}, we are
compelled to note at the outset that MCMC is in fact \emph{over-used}.
Because MCMC provably (under assumptions\note{Hint at assumptions
  here.}, some of which will be discussed below) samples the full
posterior pdf in all of parameter space, many investigators use MCMC
because it will sample \emph{all} of the parameter space
($\pars$-space).
That is, they are using MCMC because they want to
\emph{search the parameter space for good models}.
This is not a good reason to use MCMC!
Another bad use case is the following:
Because MCMC samples the parameter representatively, it spends most of
its time near very good models; models that (within the confines of
the prior pdf) do a good job of explaining the data.
For this reason, many investigators are using MCMC because it
effectively \emph{optimizes the posterior pdf}, or, for certain
choices of prior pdf, \emph{optimizes the likelihood}.\note{Of course
  a committed Bayesian would argue that any time you are optimizing a
  posterior pdf or optimizing a likelihood, you \emph{should} be
  sampling a posterior pdf.  That is, for some, the fact that when
  someone ``wants'' to optimize, it is actually useful that they
  choose the ``wrong'' tool, because that ``wrong'' tool gives them
  back something far more useful than the output of any optimizer!}
This is another bad reason!

Both of these reasons for using MCMC---that it is a parameter-space
search algorithm, and that it is a simple-to-code effective
optimizer---are \emph{not good reasons}.  MCMC is first and foremost a
\emph{sampler}.  If you are trying to find the optimum of the
likelihood or the posterior pdf, you should use an \emph{optimzer},
not a sampler.  If you want to make sure you search all of parameter
space, you should use a \emph{search algorithm}, not a sampler.  MCMC
is good at one thing, and one thing only: Sampling ill-normalized (or
otherwise hard to sample) pdfs.

In what follows, we are going to provide a kind of ``user manual'' or
advice document or folklore capture regarding the use of MCMC for
data analysis.
This will not be a detailed description of multiple MCMC methods
(indeed, we will only explain one method in detail), and it will not
be about the mathematical properties or structure of the method or
methods.
It will be about how to use MCMC, including diagnosis and
trouble-shooting.

The first couple of \sectionname s will describe what a sampling is, and how
the simplest MCMC method, the Metropolis-Hastings algorithm, can
provide one.
The next few \sectionname s will provide ideas about how to
initialize, tune and operate MCMC methods for good performance.
The last few \sectionname s will provide advice for making decisions
among the myriad MCMC methods implementations, how to implement a good
likelihood function and prior pdf function for inference, and how to
trouble-shoot standard kinds of problems that arise in operating MCMC
methods on real problems.
We will leave parenthetical and philosophical matters to the end-notes,
which appear after the main text.

@@

As we noted above (\sectionname~\ref{sec:when}), in the context of MCMC, we are often sampling or
using some \emph{badly normalized} function $f(\pars)$.
This function is just the pdf $p(\pars)$ multiplied by some unknown and
hard-to-compute scalar.
In this case, for our purposes, the conditions on $f(\pars)$ are that it
be non-negative everywhere and have \emph{finite} integral $Z$
\begin{eqnarray}
0 &\leq& f(\pars) \quad \mbox{for all $\pars$}
\\
Z &=& \int f(\pars)\,\dd \pars
\quad .
\end{eqnarray}

When the sampling $\setof{\pars_k}_{k=1}^K$ is of one of these badly
normalized functions $f(\pars)$---as it usually will be---the
sampling-approximation expectation values are the expectation values
under the properly-normalized corresponding pdf, even though you might
\emph{never learn that normalization}.
When we run MCMC sampling on $f(\pars)$, a function that differs from a
pdf $p(\pars)$ by some unknown normalization constant, then the sampling
permits computation of the following kinds of quantities:
\begin{eqnarray}
E[q] &\equiv& \frac{\int q(\pars)\,f(\pars)\,\dd \pars}{\int f(\pars)\,\dd \pars}
\\
E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
\label{eq:expectationvalue}
\quad .
\end{eqnarray}
That is, the sampling can be constructed (as we will show below) from
$f(\pars)$ directly, and it permits you to compute expectation values
without ever requiring you to integrate either the numerator integral
or the denominator integral, both of which are generally
intractable.\note{As we will
  see, the intractability comes from various directions, but one is
  that the dimension of $\pars$ gets large in most realistic situations.
  Another is that the support of $p(\pars)$ tends to be, in normal
  inference situations, \emph{far smaller} than the full domain of
  $\pars$.
  That is, the pdf is at or very close to zero over all but some tiny
  and very hard-to-find part of the space.}

The ``correctness'' of a sampling is defined (above) in terms of its use in
  performing integrals or approximate computation of integrals.
In a deep sense, the \emph{only} thing a sampling is good for is
  computing integrals.
There are many uses of the MCMC sampling, some of them good and some of them bad.
Most of the good or sensible uses will somehow involve integration.

@@

Integrals are also involved in finding the mean, median, and any quantiles
  of a pdf.
They are not involved in finding the \emph{mode} of a pdf.
For this reason (and others), in what follows, when we talk about what to report
  about the outcome of your MCMC sampling,
  we will advise in favor of mean, median, and quantiles,
  and we will advise against mode.

@@

Although the discussion in this \documentname\ is general, the most
common use of MCMC sampling (for us, anyway) is in probabilistic
inference.
For this reason, we will often refer to the function $f(\pars)$
colloquially as ``the posterior pdf''\note{We will also sometimes
  refer to expectations under $f(\pars)$ as posterior means, and
  medians as median-of-posterior values, and so on.} even though it is
implicitly ill-normalized and might not be a posterior pdf in any
sense.
We will also occasionally assume---just because it is true in
inference---that the function $f(\pars)$ is the product of two
functions, one called ``the prior pdf'' and one called ``the
likelihood''.
Again, this usage is colloquial and is only strictly correct in
inference contexts with proper inputs.
That the function $f(\pars)$ can be thought of as a product of a
prior pdf and a likelihood is only necessary for what follows in
the context of advanced sampling techniques like tempering or
nested sampling, both mentioned briefly below.

@@

The simplest algorithm for MCMC is the Metropolis--Hastings algorithm (M--H MCMC).\note{%
  There are many claims that this algorithm is incorrectly named,
  with claims that it should be credited to Enrico Fermi or Stan Ulam.
  We don't have any opinions of this matter, but encourage the reader to follow this up.
  The original paper is \citet{metropolis}, and there are a few sketchy historical notes
  in \citet{geyer}.}
It is so simple that we recommend that any reader of this document
  who has not previously implemented the algorithm take a break at the end
  of this \sectionname\ and implement it forthwith, in a short piece of computer code,
  in the context of some simple problem.\note{%
    We request this in a trivial case below in
    \problemname~\ref{prob:MH} and the following problems in this
    \sectionname.
    We make the same request in a less trivial context in our
    model-fitting \documentname\ (\citealt{fittingaline},
    \problemname~6 and 7), where we ask the Reader to implement MCMC for
    a useful mixture model.}

The M--H MCMC algorithm requires two inputs.
The first is a handle to the function $f(\pars)$ that is the function to be sampled,
  such that the algorithm can evaluate $f(\pars)$ for any value of the parameters $\pars$.
In data-analysis contexts,
  this function would be the prior $p(\pars)$ times the likelihood $p(\data\given\pars)$
  evaluated at the observed data $\data$.
The second input is a proposal pdf $q(\pars'\given\pars)$ that can be sampled,
  such that the algorithm can draw a new position $\pars'$ in the parameter space
  given an ``old'' position $\pars$.
This second function must meet a symmetry requirement (detailed
  balance) we discuss further below.
It permits us to random-walk around the parameter space in a fair way.

@@

HOGG: ...Why does M--H MCMC work (briefly)?  Here or later?

This algorithm---and indeed any MCMC algorithm---produces a
  \emph{biased random walk} through parameter space.
It is a random walk for the same reason that it is ``Markov'':
The step it makes to position $\pars_{k+1}$ depends only on the state
  of the sampler at position $\pars_k$ (and no previous state).
It is a biased random walk, biased by the acceptance algorithm
  involving the ratios of function values; this acceptance rule biases
  the random walk such that the amount of time spent in the neighborhood
  of location $\pars$ is proportional to $f(\pars)$.
Because of this local (Markov) property, the nearby samples are not
  independent; the algorithm only produces fair samples in the limit of
  arbitrary run time, and two samples are only independent when they are
  sufficiently separated in the chain (more on this below).

The principal user-settable knob in M--H MCMC is the proposal
  pdf $q(\pars'\given\pars)$.
A typical choice is a multi-variate Gaussian distribution for $\pars'$
  centered on $\pars$ with some simple (diagonal, perhaps) variance
  tensor.
We will discuss the choice and tuning of this proposal distribution
  below.

@@

As we discuss briefly below in \sectionname~\ref{sec:methods},
  there are many MCMC methods more advanced than M--H.
However, they all share characteristics with M--H
  (initialization, tuning, judging convergence),
  such that it is very valuable to understand M--H well before using anything more advanced.
Furthermore, a scientist new to MCMC benefits enormously from
  building, tuning, and using her or his own MCMC software.
One piece of advice we give then,
  to the new user of MCMC,
  is to code up, tune, and use a M--H MCMC sampler for a scientific project.
A huge amount is learned in doing this,
  and it is very often the case that the home-built M--H MCMC does everything needed;
  you often don't need any more advanced tool.
Only after you have concluded that your home-built M--H MCMC sampler
  is not suited to your project
  (or not developed properly into a properly versatile software package)
  should you download and start to use any professionally developed alternative.
That is, even if---in the end---%
  you want to leave the sampling code to the experts and run an industrial-strength code,
  it is still valuable to build your own,
  given the simplicity of the algorithm,
  and given the intuition you gain by doing it (at least once) yourself.\note{%
    HOGG Refer again to the DAR on fitting a line for an example project.}

\begin{problem}\label{prob:MH}
In your scientific programming language of choice, write a very simple
M-H MCMC sampler.
Sample in a single parameter $x$ and give the sampler as its density
function $p(x)$ a Gaussian density with mean 2 and variance 2.
Give the sampler a proposal distribution $q(x'\given x)$ a Gaussian
pdf with mean zero and variance 1.
Initialize the sampler with $x=0$ and run the sampler for more than $10^4$ steps.
Plot the results as a histogram, with the true density over-plotted sensibly.
\end{problem}

@@

\begin{problem}\label{prob:twod}
Re-do \problemname~\ref{prob:MH} but now with an input density
that is a function of two variables $(x, y)$.
For the function use a covariant Gaussian density with variance tensor HOGG,
and then a top-hat function that is uniform on HOGG and zero elsewhere.
For the proposal distribution $q(x'\given x)$ a two-dimensional Gaussian
density with zero mean and variance tensor set to the two-dimensional
identity matrix.
Plot the two one-dimensional histograms and also a two-dimensional scatter
plot for each sampling.
\end{problem}

@@

A key question for an MCMC operator---\emph{the} key question in some
sense---is \emph{how long to run} to be sure of having reliable
results.
It is disappointing and annoying to many that there is no extremely
simple and reliable answer to this question.

@@

In this context,
  it is \emph{simply not fair} to require an MCMC run to have fully and completely
  sampled the posterior pdf,
  at least not in any provable sense.
The question of whether a sampling is converged to a representative sampling
  of the posterior pdf is actually outside the domain of science,
  because the domain of science is the domain of questions answerable by scientists.
That is, you can \emph{never know} that you have correctly sampled the posterior pdf,
  despite many statements in the literature to the contrary.%
  \note{Give example statements... Mention cases in which you \emph{might} be able to claim this...}
The upshot of this is that we don't and can't require any kind of absolute convergence;
  indeed, it would be impossible even to test for it.

@@

Finally one last point about convergence:
Since all convergence tests are fundamentally heuristic, it is useful
to just make the heuristic visualization of the samples $\pars_k$ as a
function of $k$, in the order they were generated.
The chain is likely to be converged only if the random-walk process
crossed the domain of $\pars$ fully many times during the MCMC run.
Often some parameter directions are much worse than others; it is
worth looking at the chain in all parameter directions.

@@

\paragraph{Slice:}
... When is slice sampling a good idea?  DFM, can you help here?

@@

>>>>

% to-do
% -----
% - Kill all issues in this list, below:
% - Deal with all mentions of ``DFM'' or ``HOGG''
% - Go through github issues, if any.
% - Should we make an \acronym{} environment and use it religiously?  I think yes.
% - Do we need to explain the variance on a sampling-based estimator of an integral?  Right now this is just in an end-note (or will be when I write it).  Should it be in the main text?  Probably!
% - mention ESJD
% - mention Gelman--Rubin statistic
% - mention JAGS and STAN in methods section
% - follow up the idea that burn-in doesn't exist at all!  It does for emcee.
% - when we say ``below'' or ``above'' also give a Section Number.
% - need to mention case where there is an MC integral INSIDE the likelihood function
% - need to mention testing the likelihood function
% - would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
% - send to friendlies for comments
%   - Gelman
%   - Goodman
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
% - post on arXiv

% style notes
% -----------
% - \MCMC not MCMC
% - These documents are ``Chapters'', no?  Use \documentname.
% - [LaTeX] newline after every full stop for proper git diffing
% - [LaTeX] eqnarray not equation
% - pdf not PDF
% - is it ``a MCMC sampling'' or ``an MCMC sampling''?
% - \note{} right after word if you are endnoting a phrase,
%   but after the full stop if endnoting the sentence or paragraph.
% - always $\pars$ never $x$; $K$ samples $\pars_k$ and $\pars$ space is $D$-dimensional
% - should we use the word ``walker'' everywhere and do we?

@@

\newcommand{\MCMC}{\acronym{MCMC}}
\newcommand{\MH}{\acronym{M--H}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\data}{D}
\newcommand{\pars}{\theta}
\newcommand{\best}{{\mathrm{(best)}}}
\newcommand{\better}{{\mathrm{(better)}}}

@@

\begin{abstract}
Markov Chain Monte Carlo (\MCMC) methods for sampling probability
distribution functions (combined with abundant computational resources)
have transformed the sciences.
Here we give a brief overview of the most basic \MCMC\ method and then turn to
practical advice for the use of \MCMC\ in real inference problems.
We give advice on method choice, tuning for performance,
methods for initialization, tests of convergence, troubleshooting, and use of the chain
output to produce or report parameter estimates with associated uncertainties.
\end{abstract}

@@

Markov Chain Monte Carlo (\MCMC) methods are methods for sampling
probability distribution functions or probability density functions (pdfs).
These pdfs may be either probability mass functions on a discrete
space, or probability densities on a continuous space, though we will
concentrate on the latter in this \documentname.
\MCMC\ methods don't require that you have a full analytic description of the
properly normalized pdf for sampling to proceed; they only require
that you be able to compute ratios of the pdf at pairs of locations.
This makes \MCMC\ methods ideal for sampling \emph{posterior
  pdfs} in probabilistic inferences:

@@

In addition to this normalization-insensitive property of \MCMC, in its
simplest forms it can be run without computing any derivatives or
integrals of the function, and (as we will show below) in its simplest
forms it is \emph{extremely easy to implement}.
For all these reasons, \MCMC\ is ideal for sampling posterior pdfs in
the real situations in which scientists find themselves.

Say you are in this situation:
You have a huge blob of data $\data$ (think of this as a vector or
list or heterogeneous collection of observations\note{In general in
  this \documentname\ we will treat any variable as being capable of
  containing not just a value but some huge arbitrarily structured set
  of values in matrix, vector, or an ordered list of any set of such objects,
  or perhaps some even more complex form.}).
You also have a model sophisticated enough---a probabilistic,
generative model, if you will\note{For more discussion
  of our definition of a \emph{model}, see [some other as-yet-unwritten \documentname] in
  this series.
  Briefly, a ``model'' for us is a likelihood function (a pdf for the data given
  model parameters), and a prior pdf over the parameters.
  Because, under this definition, the model can always generate (by
  sampling, say) parameters and parameters can generate (again by
  sampling, say) data, the model is effectively (or actually, if you
  are a true subjective Bayesian) a probability distribution (pdf)
  over all possible data.%
}---that, given a
setting of a huge blob of parameters\note{Again, the parameter
  blob $\pars$ should also be thought of as an immense, arbitrarily
  structured set or list of parameters, possibly even with no fixed
  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We will
  bash the terminology ``likelihood'' in another, as-yet-unwritten, \documentname\ in this
  series.  Techically, this thing $p(\data\given\pars)$
  is only properly a likelihood function when
  we are thinking of the data $\data$ as being fixed, and the parameters $\pars$
  as being permitted
  to vary.}) $p(\data\given\pars)$.
Furthermore, say also that you can write down some kind of informative or
vague prior pdf $p(\pars)$ for the parameter blob $\pars$.
If all these things are true, then---even if you can't compute
anything else---in principle a trivial-to-implement \MCMC\ can give you
a fair sampling of the posterior pdf.
That is, you can run \MCMC\ 
(for a very long time---see \sectionname~\ref{sec:convergence} for how long)
and you will be left with a set of $K$ parameter-blob settings $\pars_k$ such that the
full set $\setof{\pars_k}_{k=1}^K$ constitutes a fair sampling from
the posterior pdf $p(\pars\given\data$).
% ************* BJB
% Notions of "sampling" and "Monte Carlo" are interesting. You might like to
% refer the interested reader to the paper "Monte Carlo is Fundamentally
% Unsound" by Tony O'Hagan.

All that said, and adhering to the traditions of the \project{Data
  Analysis Recipes} project\note{Every \documentname\ in the \project{Data
    Analysis Recipes} series begins with a rant in which we argue that
  most uses of the methods in question are not appropriate!}, we are
compelled to note at the outset that \MCMC\ is in fact \emph{over-used}.
Because \MCMC\ provably (under assumptions\note{Hint at assumptions
  here.}, some of which will be discussed below) samples the full
posterior pdf in all of parameter space, many investigators use \MCMC\ 
because it will sample \emph{all} of the parameter space
($\pars$-space).
That is, they are using \MCMC\ because they want to
\emph{search the parameter space for good models}.
This is not a good reason to use \MCMC!
Another bad use case is the following:
Because \MCMC\ samples the parameter representatively, it spends most of
its time near very good models; models that (within the confines of
the prior pdf) do a good job of explaining the data.
For this reason, many investigators are using \MCMC\ because it
effectively \emph{optimizes the posterior pdf}, or, for certain
choices of prior pdf, \emph{optimizes the likelihood}.\note{Of course
  a committed Bayesian would argue that any time you are optimizing a
  posterior pdf or optimizing a likelihood, you \emph{should} be
  sampling a posterior pdf.  That is, for some, the fact that when
  someone ``wants'' to optimize, it is actually useful that they
  choose the ``wrong'' tool, because that ``wrong'' tool gives them
  back something far more useful than the output of any optimizer!}
This is another bad reason!

Both of these reasons for using \MCMC---that it is a parameter-space
search algorithm, and that it is a simple-to-code effective
optimizer---are \emph{not good reasons}.  \MCMC\ is first and foremost a
\emph{sampler}.  If you are trying to find the optimum of the
likelihood or the posterior pdf, you should use an \emph{optimzer},
not a sampler.  If you want to make sure you search all of parameter
space, you should use a \emph{search algorithm}, not a sampler.  \MCMC\ 
is good at one thing, and one thing only: Sampling ill-normalized (or
otherwise hard to sample) pdfs.

In what follows, we are going to provide a kind of ``user manual'' or
advice document or folklore capture regarding the use of \MCMC\ for
data analysis.
This will not be a detailed description of multiple \MCMC\ methods
(indeed, we will only explain one method in detail), and it will not
be about the mathematical properties or structure of the method or
methods.
It will be about how to use \MCMC, including diagnosis and
trouble-shooting.

The first couple of \sectionname s will describe what a sampling is, and how
the simplest \MCMC\ method, the Metropolis-Hastings algorithm, can
provide one.
The next few \sectionname s will provide ideas about how to
initialize, tune and operate \MCMC\ methods for good performance.
The last few \sectionname s will provide advice for making decisions
among the myriad \MCMC\ methods implementations, how to implement a good
likelihood function and prior pdf function for inference, and how to
trouble-shoot standard kinds of problems that arise in operating \MCMC\ 
methods on real problems.
We will leave parenthetical and philosophical matters to the end-notes,
which appear after the main text.

@@

As we noted above (\sectionname~\ref{sec:when}), in the context of \MCMC, we are often sampling or
using some \emph{badly normalized} function $f(\pars)$.
This function is just the pdf $p(\pars)$ multiplied by some unknown and
hard-to-compute scalar.
In this case, for our purposes, the conditions on $f(\pars)$ are that it
be non-negative everywhere and have \emph{finite} integral $Z$
\begin{eqnarray}
0 &\leq& f(\pars) \quad \mbox{for all $\pars$}
\\
Z &=& \int f(\pars)\,\dd \pars
\quad .
\end{eqnarray}

When the sampling $\setof{\pars_k}_{k=1}^K$ is of one of these badly
normalized functions $f(\pars)$---as it usually will be---the
sampling-approximation expectation values are the expectation values
under the properly-normalized corresponding pdf, even though you might
\emph{never learn that normalization}.
When we run \MCMC\ sampling on $f(\pars)$, a function that differs from a
pdf $p(\pars)$ by some unknown normalization constant, then the sampling
permits computation of the following kinds of quantities:
\begin{eqnarray}
E[q] &\equiv& \frac{\int q(\pars)\,f(\pars)\,\dd \pars}{\int f(\pars)\,\dd \pars}
\\
E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
\label{eq:expectationvalue}
\quad .
\end{eqnarray}
That is, the sampling can be constructed (as we will show below) from
$f(\pars)$ directly, and it permits you to compute expectation values
without ever requiring you to integrate either the numerator integral
or the denominator integral, both of which are generally
intractable.\note{As we will
  see, the intractability comes from various directions, but one is
  that the dimension of $\pars$ gets large in most realistic situations.
  Another is that the support of $p(\pars)$ tends to be, in normal
  inference situations, \emph{far smaller} than the full domain of
  $\pars$.
  That is, the pdf is at or very close to zero over all but some tiny
  and very hard-to-find part of the space.}

The ``correctness'' of a sampling is defined (above) in terms of its use in
  performing integrals or approximate computation of integrals.
In a deep sense, the \emph{only} thing a sampling is good for is
  computing integrals.
There are many uses of the \MCMC\ sampling, some of them good and some of them bad.
Most of the good or sensible uses will somehow involve integration.

@@

Integrals are also involved in finding the mean, median, and any quantiles
  of a pdf.
They are not involved in finding the \emph{mode} of a pdf.
For this reason (and others), in what follows, when we talk about what to report
  about the outcome of your \MCMC\ sampling,
  we will advise in favor of mean, median, and quantiles,
  and we will advise against mode.

@@

Although the discussion in this \documentname\ is general, the most
common use of \MCMC\ sampling (for us, anyway) is in probabilistic
inference.
For this reason, we will often refer to the function $f(\pars)$
colloquially as ``the posterior pdf''\note{We will also sometimes
  refer to expectations under $f(\pars)$ as posterior means, and
  medians as median-of-posterior values, and so on.} even though it is
implicitly ill-normalized and might not be a posterior pdf in any
sense.
We will also occasionally assume---just because it is true in
inference---that the function $f(\pars)$ is the product of two
functions, one called ``the prior pdf'' and one called ``the
likelihood''.
Again, this usage is colloquial and is only strictly correct in
inference contexts with proper inputs.
That the function $f(\pars)$ can be thought of as a product of a
prior pdf and a likelihood is only necessary for what follows in
the context of advanced sampling techniques like tempering or
nested sampling, both mentioned briefly below.

@@

The simplest algorithm for \MCMC\ is the Metropolis--Hastings algorithm (\MH\ \MCMC).\note{%
  There are many claims that this algorithm is incorrectly named,
  with claims that it should be credited to Enrico Fermi or Stan Ulam.
  We don't have any opinions of this matter, but encourage the reader to follow this up.
  The original paper is \citet{metropolis}, and there are a few sketchy historical notes
  in \citet{geyer}.}
It is so simple that we recommend that any reader of this document
  who has not previously implemented the algorithm take a break at the end
  of this \sectionname\ and implement it forthwith, in a short piece of computer code,
  in the context of some simple problem.\note{%
    We request this in a trivial case below in
    \problemname~\ref{prob:MH} and the following problems in this
    \sectionname.
    We make the same request in a less trivial context in our
    model-fitting \documentname\ (\citealt{fittingaline},
    \problemname~6 and 7), where we ask the Reader to implement \MCMC\ for
    a useful mixture model.}

The \MH\ \MCMC\ algorithm requires two inputs.
The first is a handle to the function $f(\pars)$ that is the function to be sampled,
  such that the algorithm can evaluate $f(\pars)$ for any value of the parameters $\pars$.
In data-analysis contexts,
  this function would be the prior $p(\pars)$ times the likelihood $p(\data\given\pars)$
  evaluated at the observed data $\data$.
The second input is a proposal pdf $q(\pars'\given\pars)$ that can be sampled,
  such that the algorithm can draw a new position $\pars'$ in the parameter space
  given an ``old'' position $\pars$.
This second function must meet a symmetry requirement (detailed
  balance) we discuss further below.
It permits us to random-walk around the parameter space in a fair way.

@@

HOGG: ...Why does \MH\ \MCMC\ work (briefly)?  Here or later?

This algorithm---and indeed any \MCMC\ algorithm---produces a
  \emph{biased random walk} through parameter space.
It is a random walk for the same reason that it is ``Markov'':
The step it makes to position $\pars_{k+1}$ depends only on the state
  of the sampler at position $\pars_k$ (and no previous state).
It is a biased random walk, biased by the acceptance algorithm
  involving the ratios of function values; this acceptance rule biases
  the random walk such that the amount of time spent in the neighborhood
  of location $\pars$ is proportional to $f(\pars)$.
Because of this local (Markov) property, the nearby samples are not
  independent; the algorithm only produces fair samples in the limit of
  arbitrary run time, and two samples are only independent when they are
  sufficiently separated in the chain (more on this below).

The principal user-settable knob in \MH\ \MCMC\ is the proposal
  pdf $q(\pars'\given\pars)$.
A typical choice is a multi-variate Gaussian distribution for $\pars'$
  centered on $\pars$ with some simple (diagonal, perhaps) variance
  tensor.
We will discuss the choice and tuning of this proposal distribution
  below.

@@

As we discuss briefly below in \sectionname~\ref{sec:methods},
  there are many \MCMC\ methods more advanced than M--H.
However, they all share characteristics with M--H
  (initialization, tuning, judging convergence),
  such that it is very valuable to understand \MH\ well before using anything more advanced.
Furthermore, a scientist new to \MCMC\ benefits enormously from
  building, tuning, and using her or his own \MCMC\ software.
One piece of advice we give then,
  to the new user of \MCMC,
  is to code up, tune, and use a \MH\ \MCMC\ sampler for a scientific project.
A huge amount is learned in doing this,
  and it is very often the case that the home-built \MH\ \MCMC\ does everything needed;
  you often don't need any more advanced tool.
Only after you have concluded that your home-built \MH\ \MCMC\ sampler
  is not suited to your project
  (or not developed properly into a properly versatile software package)
  should you download and start to use any professionally developed alternative.
That is, even if---in the end---%
  you want to leave the sampling code to the experts and run an industrial-strength code,
  it is still valuable to build your own,
  given the simplicity of the algorithm,
  and given the intuition you gain by doing it (at least once) yourself.\note{%
    We also make this point in a previous \documentname\ (\citealt{fittingaline};
    the ambitious reader will find in that \documentname\ a solid example project
    for using \MCMC\ in a real context.}

\begin{problem}\label{prob:MH}
In your scientific programming language of choice, write a very simple
M-H \MCMC\ sampler.
Sample in a single parameter $x$ and give the sampler as its density
function $p(x)$ a Gaussian density with mean 2 and variance 2.
(Note that variance is the \emph{square} of the standard deviation.)
Give the sampler a proposal distribution $q(x'\given x)$ a Gaussian
pdf with mean zero and variance 1.
Initialize the sampler with $x=0$ and run the sampler for more than $10^4$ steps.
Plot the results as a histogram, with the true density over-plotted sensibly.
\end{problem}

@@

\begin{problem}\label{prob:twod}
Re-do \problemname~\ref{prob:MH} but now with an input density
that is a function of two variables $(x, y)$.
For the density function use two different functions.
The first density function is a covariant two-dimensional Gaussian density with variance tensor
\begin{eqnarray}
V &=& \left[\begin{array}{cc} 2.0 & 1.2 \\ 1.2 & 2.0 \end{array}\right]
\quad .
\end{eqnarray}
The second density function is a rectangular top-hat function that is uniform on
the joint constraint $3<x<7$ and $1<y<9$ and zero everywhere else.
For the proposal distribution $q(x', y'\given x, y)$ a two-dimensional Gaussian
density with zero mean and variance tensor set to the two-dimensional
identity matrix.
Plot the two one-dimensional histograms and also a two-dimensional scatter
plot for each sampling.
\end{problem}

@@

A key question for an \MCMC\ operator---\emph{the} key question in some
sense---is \emph{how long to run} to be sure of having reliable
results.
It is disappointing and annoying to many that there is no extremely
simple and reliable answer to this question.

@@

In this context,
  it is \emph{simply not fair} to require an \MCMC\ run to have fully and completely
  sampled the posterior pdf,
  at least not in any provable sense.
The question of whether a sampling is converged to a representative sampling
  of the posterior pdf is actually outside the domain of science,
  because the domain of science is the domain of questions answerable by scientists.
That is, you can \emph{never know} that you have correctly sampled the posterior pdf,
  despite many statements in the literature to the contrary.%
  \note{Give example statements... HOGG: Mention cases in which you \emph{might} be able to claim this...}
The upshot of this is that we don't and can't require any kind of absolute convergence;
  indeed, it would be impossible even to test for it.

@@

Finally one last point about convergence:
Since all convergence tests are fundamentally heuristic, it is useful
to just make the heuristic visualization of the samples $\pars_k$ as a
function of $k$, in the order they were generated.
The chain is likely to be converged only if the random-walk process
crossed the domain of $\pars$ fully many times during the \MCMC\ run.
Often some parameter directions are much worse than others; it is
worth looking at the chain in all parameter directions.

@@

\paragraph{Slice:}
DFM ... When is slice sampling a good idea?  DFM, can you help here?

@@

\begin{problem}
For a real-world inference problem, read enough of
\citet{fittingaline} to understand and execute Exercise~6 in that
document.
\end{problem}
